Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-03-11 text/front_end_v3.qmd
revised front end

On January 9, 2019 a public research university purchased three "student lists" from College Board in order to obtain the contact information of out-of-state prospects, who would subsequently be sent email, mail, and targeted social media. Each purchase utilized three "search filters," thereby controlling which prospects profiles were purchased. First, all three student list purchases filtered for the 2020 high school class. Second, under the heading "College Board Exams," each purchase filtered on a different PSAT score range: 1070 - 1180 for order #448922; 1190 - 1260 for order #448427; and 1270 - 1520 for order #448440. Third, under the heading "Geography", each purchase filtered on the same 67 "Geomarkets," which had titles like "IL07 - Chain of Lakes", "IL08 - Northwest Suburbs", "IL09 - North Shore."

Geomarkets carve states and metroplitan areas into smaller geographic regions meant to define local recruiting markets. A closer look at the Geomarkets selected for orders #448922, #448427, and #448440 shows that each order selected all Geomarkets in the Chicago-land area, except "IL10 - City of Chicago", which is significantly poorer and less White than surrounding Geomarkets. Similarly, in the DC/Maryland/Virginia metro area, each order included the "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but not the "MD05 - Prince Georges Metropolitan" "MD07 - Baltimore (Urban)" or "DC01 - District of Columbia" Geomarkets which are poorer and more Black. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. However, the above Geomarket selections raise the concern that Geomarkets -- and other geographic search filters -- may be means of excluding working class and non-white students from educational opportunities.

This manuscript does blah blah.

In 1983, Robert Zemsky and Penney Oedel authored *The Structure of College Choice*. Based on an analysis of 1980 SAT score-sending behavior and published by *The College Board*, @RN4982 developed the Market Segment Model to help colleges recruit students. The underlying idea of the Market Segment Model is that student demand for your institution is fundamentally a function of class. Therefore, the key to recruitment is identifying territories where large numbers of your class live. @RN4982 [p. 11] describe the Market Segment Model as an effort to "quantify" the knowledge that college admissions officers have about the geography of student demand:

> A good recruiter knows where to look for prospective applicants, as seen in the students' willingness or eluctance to travel. It is necessary to identify not only the most promising communities or poools, but also the specific neighborhoods within those communities -- hence the recruiter's emphasis on feeder high schools. From the beginning, we have sought to mold our research to these concepts, to capture and quantify the phenomenon behind the folklore. Thus, it is the admissions officers' notion of admissions pools that dictates the geographic units of our analysis." \[p. 11\]

@RN4982 also created local "community based markets" (p. 14) -- later known as "Geomarkets." Geomarkets are geographic borders that disaggregate states and large metropolitan areas for recruiting purposes. @fig-nyc-geomarkets shows New York City area Geomarkets. In defining Geomarkets, @RN4982 did not draw new geographic borders, but rather made certain extant borders more salient to the process of recruiting.

Geomarkets have been incorporated into the supply-side structure of college access in ways that are simultaneusly mundane and fundamental. We posit three ways. First, Geomarkets became a principle for how many -- if -- admissions offices organized their labor. In states where a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory \[CITE\]. Geomarkets were created to mirror the territories of admissions officers and, in turn, admissions offices structured their territories around Geomarkets. Second, The Market Segment Model and Geomarkets became the basis for Geomarkets are the basis for the College Board Enrollment Planning Service (EPS), founded in 1984 and still active today. EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets. @RN5019_2010 says "Enrollment Planning Service (EPS\textsuperscript{\textregistered}) the analysis tool that pinpoints the schools and geomarkets where your best prospects are most likely to be found." @RN5021 reports that, in 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS.

Third, Geomarkets were incorporated into the College Board "student list" product named Student Search Service. Student lists have been the primary source of lead generation in U.S. higher education since 1972, when College Board began selling "names" [@belkin2019-studata; @list_biz]. <!-- @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. Lists contain contact information for prospective students.--> The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50 in 2021). Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular Geomarkets. An analysis of 830 student lists purchased by 14 public universities found that X% of purchases filtered on Geomarket [@RN5012].

Geomarkets have never been the subject of empirical research. This omission is surprising because the sociology of education is concerned with sorting [@RN4873], particularly sorting into college. Traditionally, most scholarship -- and certainly most scholarship in economics \[HOXBY? CHETTY\] -- views inequality in college access as a function households or k-12 schools, which represent the demand-side of higher education. More recently, scholarship in the sociology of education takes seriously the idea that colleges are not passive recipients of applications, but rather they expend substantial resources soliciting demand from desirable applicants [@RN3519; @RN4774; @RN4324]. However, most scholarship on recruiting analyzes the behavior of individual colleges [e.g., @RN4758; @RN4759], thereby creating the assumption that recruiting is a function of individual colleges. What these studies have missed is the broader enrollment management industry -- third-party vendors, their products and consulting solutions -- as a set of mechanisms that structure college access, funneling certain kinds of students to certain kinds of institutions.

This manuscript analyzes one of those mechanisms, The Market Segment Model and College Board Geomarkets, as a case study of quantification. We develop arguments from the broader literature on quantification [e.g., @RN5003; @RN5005; @RN5007], particularly the discussions of correlation and homophily from @RN4975. Predictive analytics are based on correlation. When we use data on past correlations to make recommendations about the future, we amplify the effects of historic structural inequality. The Market Segment Model analyzes SAT score-sending data from 1980 and concludes that student demand for higher education is primarily a function of social class [@RN4982]. The Market Segment Model argues that homophily -- actors that share characteristics are likely to form connections -- is the organizing principle of student demand and competition between colleges. @RN4982 [p. 72] state that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students." This snapshot of student demand in 1980, itself a consequence historic structural inequality, was programmed into the way admissions offices organize labor and programmed into the EPS and Student Service Service products that colleges utilize to identify and target prospective students. The result is a supply-side that reinforces structural inequalities observed on the demand-side.

\[ADD TEXT SOMEWHERE AROUND HERE ABOUT WHY THIS TOPIC IS SIGNIFICANT AND IMPORTANT FOR CURRENT/FUTURE POLICY DEBATES: Why market segment model important. Software as service products dominate em industry and structure college access. Look at eab machine learning product But rarely do we see the recipe underneath these products. The book structuring college access is that recipe for EPS software\]

Our analyses address the two research questions, which speak to how geomarkets are utilized within EPS and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining the geomarket shapefile to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular geomarkets? We address this question by analyzing actual student list purchases that utilize commonly used search filters (e.g., PSAT score, GPA) but do not filter on Geomarkets. We simulate which prospects in particular metropolitan areas would have been included/excluded had the student list purchase filtered on particular Geomarkets. We obtained these data by issuing public records requests to public universities. <!-- Third, in student list purchases that filtered on geomarkets, how does the socioeconomic and racial composition of included students compare the that of the geomarket and how does it compare to that of the broader metropolitan area?  -->

```{=html}
<!-- 
THIS TEXT IS FINE, BUT TRYING TO SET UP THE CONTRIBUTION TO BROADER QUANTIFICATION LITERATURE, VIS-A-VIS NO-ONE HAS STUDIED MARKET RESEARCH; DON'T THINK THIS IS NECESSARY AND YOU ARE NOT ON FIRM GROUND HERE IN CALLING THIS "MARKET RESEARCH"
Scholarship on quantification in education has focused on consumer-facing measures such as U.S. News Rankings [@RN5003] and UK "league tables" about school quality [@RN5005]. To our knowledge, prior knowledge has not examined business-facing measures produced by third-party market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students. We reason that market research reproduces inequality by encouraging colleges to plan their recruiting efforts based on a snapshot of social stratification in student demand that is substantially a function of historic, structural inequality in educational opportunity. In contrast to @RN5005, we cannot demonstrate the effect of geomarkets on inequality. However, we show the mechanisms of how geomarkets reproduce inequality by simulating who is included/excluded when geomarkets are utilized within student list purchases as recommended by @RN4982.
-->
```

In the following section we provide background information about enrollment management and scholarship on recruiting. Second, we introduce core ideas and concepts from scholarship on quantification. Third, we apply these ideas to a close read of @RN4982 and motivate hypotheses. Four, we describe data and methods. Fifth, we present results. Finally, we discuss implications for scholarship and for policy.

A note from RESULTS that I texted to Bastedo on 12/19/2024

-   But these geomarkets are pretty fucked up. They were designed to capture class. I see lots of orders that condition on test scores and rich geomarkets. But I know from similar purchases that don’t condition on geomarket, that the high scoring first gen Asian and Hispanic students tend to come from the poor geomarkets that get excluded from a lot of purchases. By contrast, The Asian and Hispanic students that do get included in purchases that target rich geomarkets are usually not first gen

@@

@fig-em-funnel depicts the "enrollment funnel," which modifies the traditional "marketing funnel" to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution, consisting of those that respond to an initial soliciation (e.g., email) and those that reach out on their own (e.g., sending SAT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying. The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply).  -->The purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media advertisements designed to solicit inquiries and applications [@RN4895].

@@

Drawing from scholarship on organizational theory [@RN4738], enrollment management processes involve many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].  In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].

@@

**Correlation**. The interdisciplinary literature on quantification [@RN5008] includes important contributions from the field of critical data studies [e.g., @RN4849; @RN4772; ]. In particular, the discussions of correlation (chapter 1) and homophily (chapter 2) by @RN4975 introduce ideas salient to the analysis the Market Segment Model and College Board Geomarkets. Correlation measures the extent to which two or more variables move together. Predictive analytics are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \ldots, big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"^[Due to data limitations, predictive analytics often utilize "proxy" variables [@RN4849; @RN4975], which are variables that are correlated with a variable of interest. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?']

@@

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Like correlational models discussed above, network science models "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. <!--"Homophily assumes and creates segregation" [@RN4975, p. 95] writes. "It transforms individuals into 'neighbors' who naturally want to live with people 'like them'." --> In commercial social networks, @RN4975 argues, homopphily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "social networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@RN4975, p. YY]. We observe similar processes in college access because higher education is an industry in which providers are evaluated based on the characteristics of their customers [@RN1549; @RN1237]. Based on a snapshot of 1980 SAT score-sending, @RN4982 concludes that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reasons that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side -- which is a function of historic structural inequality –- is programmed into the supply-side.

@@

Homophily is also central to products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].[^4] Classification situations engender markets where a vertical hierarchy of products are targeted to a vertical hierarchy of consumers. We observe similar processes in our case study, whereby @RN4982 categorizes students into four market segments -- local, in-state, regional, and national -- based on their score sending behavior and then evaluates the attractiveness of each Geomarket based on how many "regional" and "national" students live there.
<!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables, @RN5005 we cannot show the effect of quantification. However, this study provides inisight about the mechanism underlying the effect. By analyzing and simulating *Student Search Service* purchases that filter on Geomarkets we show how Geomarkets reproduce historical race-based inequality in college access [THIS LAST BIT FEELS LIKE THE GENERAL CONTRIBUTION TO SOCIOLOGY].

@@

**Creating Geomarkets and the Market Segment Model.** In 1978, University of Pennsylvania professor Robert Zemsky, was tasked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x].^[Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s.] Zemsky reached out to College Board because "we needed a database that described most institutions and most students\ldots Coincidentally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential" [@RN4982, p. x].

The geographic concepts inscribed in the Market Segment Model were based on conversations with college admissions officers. @RN4982 \[p. 4\] state that "our initial task was to define enrollment markets in a manner consistent with admissions officers' intuitive understanding of student pools." Quoting an admissions officer, @RN4982 [p. 11] write, "'There are only three kinds of colege-bound students: those who want to live at home, those who want to live on campus but bring their laundry home, and those who want to go far enough from home that Mom and Dad can't visit without calling first.'" As such, [@RN4982] created three geographic [@RN4982, p. 11], region, state, and community. The three regions were New England, Middle States, and the South. Next, "we divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets." These enrollment markets, later called Geomarkets, were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. @RN4982[pp. 11-12] briefly describes the creation of Geomarket borders:

@@

The goal of the Market Segment Model was to quantify the knowledge that admissions counselors have about student demand. In 1979, College Board began providing funding and data on the score-sending behavior of SAT test-takers. @RN4982 [p. 4] write that,

> For our analysis, we sought not a complex mathematical model, but a straightforward classification system that would track the pattern of SAT-score submissions to create a map of student choice. The Market Segment model that we developed was nothing more than a set of simple rules for disaggregating high school seniors into similar groups. The model worked because students, once so disaggregated, appeared to behave in remarkably consistent ways.

>>>>

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased. First, the order filtered on 2020 high school graduating class. Second, under the heading "College Board Exams," the order filtered on PSAT score range of 1070 - 1180. Third, under the heading "Geography", the order filtered on 67 "Geomarkets." Geomarkets carve states and metroplitan areas into smaller geographic units and have names like, IL08 - Northwest Suburbs" and "IL09 - North Shore."

On the same day, the university also purchased orders $\#448427$ ($57,537$ names) and $\#448440$ ($78,000$ names), which filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets as order $\#448922$. On closer inspection, the orders filtered on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White than surrounding Geomarkets. In the DC/Maryland/Virginia metro area, the orders included the "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 [check] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.

The origins of Geomarkets, and how they were incorporated into recruiting products, turns out to be an interesting story. In 1983, Robert Zemsky and Penny Odell authored *The Structure of College Choice*. College Board published the book, provided data, and financed the project, as part of their "efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest geographic unit used in these analyses was the state" [@RN4982, p. x]. @RN4982 sought to "capture and quantify" (p. 11) the knowledge of admissions officers,

> Admissions officers invariably are tellers of stories -- about the colleges they represent, about the colleges they attended, about each other, and about the often vagabond life of college recruiting (p.9)\ldots We have begun with this celebration of sotrytelling\ldots [because] we believe that the intuitios of admissions officers actually comprise a remarkably systematic body of knowledge about the college selection process\ldots Our research this is based opn listening carefully to what admissions officers have to say (pp. 9-10).

@RN4982 developed Geomarkets and the Market Segment Model based on the idea that "a good recruiter knows where to look for prospective applicants, as seen in the students' willingness or reluctance to travel." Created by analyzing the SAT score-sending behavior of 1980 high school seniors, the thesis of the Market Segment Model is that student demand for your institution is a function of social class and geography. Therefore, colleges should recruit from territories that contain large populations of your target social class. Geomarkets are geographic borders meant to define local recruiting markets, a territory of an admissions recruiter. Figure X, panel A shows Geomarkets in the Chicago-land area. The Market Segment Model predicts how demand for a particular college varies by Geomarket, based on characteristics of households in the Geomarket.

Geomarkets have been incorporated into the supply-side structure of college access in several ways. Albeit anecdotally, Geomarkets became an organizing principle for how college admissions offices allocate admissions recruiters to territories. Figure X panel B shows that Chicago-land Geomarkets define recruiting territories for University of Chicago admissions recruiters (Appendix Figures BLAH BLAH). Second, Geomarkets are the basis for the College Board Enrollment Planning Service (EPS). Founded in 1984 and still active today, EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets. According to @RN5021, in 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis product, which was based on EPS.

Third, Geomarkets were incorporated into College Board's student list product, Student Search Service. Student lists contain the contact information of prospective students and have been the primary source of lead generation in U.S. higher education since 1972, when College Board began selling names [@belkin2019-studata; @list_biz]. @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50 in 2021). Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular Geomarkets.

Scholarship at the nexus of college access and recruiting has focused on the behavior of colleges [e.g., @cottom2017lower; @RN4758; @RN4759; @RN3519], contributing to the belief that recruiting is a function of individual colleges. Scholarship has failed to investigate the broader enrollment management industry -- consisting of third-party vendors, their products and consulting solutions -- as a set of mechanisms that structure college access. These mechanisms, which lie upstream of the behavior of individual colleges, are incorporated by individual colleges in ways that funnel certain kinds of students to certain kinds of institutions. This omission is surprising because sociology is concerned with sorting [@RN4873], particularly sorting into college.

This manuscript analyzes College Board Geomarkets as a case study of quantification. In particular, we draw from the discussions of correlation and homophily from @RN4975. @RN4982 identified the correlates of 1980 SAT score-sending behavior and concluded that student demand for higher education is primarily a function of social class. The Market Segment Model argues that homophily -- actors that share characteristics form connections -- is the organizing principle of competition and college choice, stating that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students" [@RN4982, p. 72]. Scholarship on quantification demonstrates that making recommendations based on past correlations amplifies the effects of historic structural inequality [@RN4794]. The snapshot of student demand in 1980 -- itself a consequence historic structural inequality -- was programmed into recruiting products that colleges utilize to identify and target prospective students. The result is a supply-side that amplifies structural inequalities observed on the demand-side.

Our analyses address the two research questions, which speak to how Geomarkets are utilized within EPS and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining the geomarket shapefile to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular geomarkets? We address this question by analyzing actual student list purchases that utilize commonly used search filters (e.g., PSAT score, GPA) but do not filter on Geomarkets. We simulate which prospects in particular metropolitan areas would have been included/excluded had the student list purchase filtered on particular Geomarkets. We obtained these data by issuing public records requests to public universities.

The following section provides background about enrollment management and introduces salient concepts from scholarship on quantification. Second, we introduce the case, describing how @RN4982 developed Geomarkets and the Market Segment Model and how these concepts were incorporated into College Board recruiting products. Third, we describe data and methods. Fourth, we present results. [We find that\ldots ?PARAGRAPH ON RESULTS?] Finally, we discuss implications for scholarship and for policy. The sociology of education should move beyond the fixation on schools and policies as the primary sorting machines. @RN527[p. 148] remind us that the organizational field includes the totality of relevant actors." Third-party vendors have structured college access for decades and their influence is growing. The contemporary enrollment management industry is characterized by private equity owned firms selling software-as-service platforms that utilize the same logic as @RN4982, for example EAB's "pipeline analytics" machine learning product [CITE]. By investigating high-leverage actors and products in the organizational field, sociology can develop important insights that shape future federal and state regulatory policy.

```{=html}
\[ADD TEXT SOMEWHERE AROUND HERE ABOUT WHY THIS TOPIC IS SIGNIFICANT AND IMPORTANT FOR CURRENT/FUTURE POLICY DEBATES: Why market segment model important. Software as service products dominate em industry and structure college access. Look at eab machine learning product But rarely do we see the recipe underneath these products. The book structuring college access is that recipe for EPS software\]
<!-- 
THIS TEXT IS FINE, BUT TRYING TO SET UP THE CONTRIBUTION TO BROADER QUANTIFICATION LITERATURE, VIS-A-VIS NO-ONE HAS STUDIED MARKET RESEARCH; DON'T THINK THIS IS NECESSARY AND YOU ARE NOT ON FIRM GROUND HERE IN CALLING THIS "MARKET RESEARCH"
Scholarship on quantification in education has focused on consumer-facing measures such as U.S. News Rankings [@RN5003] and UK "league tables" about school quality [@RN5005]. To our knowledge, prior knowledge has not examined business-facing measures produced by third-party market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students. We reason that market research reproduces inequality by encouraging colleges to plan their recruiting efforts based on a snapshot of social stratification in student demand that is substantially a function of historic, structural inequality in educational opportunity. In contrast to @RN5005, we cannot demonstrate the effect of geomarkets on inequality. However, we show the mechanisms of how geomarkets reproduce inequality by simulating who is included/excluded when geomarkets are utilized within student list purchases as recommended by @RN4982.
-->
```

@@

@fig-em-funnel depicts the "enrollment funnel," which modifies the traditional "marketing funnel" to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution, including those that respond to an initial solicitation (e.g., email) and those that reach out on their own (e.g., sending SAT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying. The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply).  -->The purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media designed to solicit inquiries and applications [@RN4895].

@@

Drawing from scholarship on organizational theory, enrollment management processes involve many "make or buy" [@RN4738] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].  In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].

@@

**Correlation**. The interdisciplinary literature on quantification [@RN5008] includes important contributions from the field of critical data studies [e.g., @RN4772; @RN4849]. In particular, the discussions of correlation (chapter 1) and homophily (chapter 2) by @RN4975 introduce ideas salient to the analysis the Market Segment Model and Geomarkets. Correlation measures the extent to which two or more variables move together. Predictive analytics are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \ldots, big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"^[Due to data limitations, predictive analytics often utilize "proxy" variables [@RN4849; @RN4975], which are variables that are correlated with a variable of interest. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?']

@@

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Like correlational models discussed above, because most network science models do not observe how historical structures affect relationships, they "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. Second, in commercial social networks, homopphily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "social networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@RN4975, p. YY]. We observe similar processes in college access. Based on a snapshot of 1980 SAT score-sending, @RN4982 concludes that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reasons that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side -- which is a function of historic structural inequality –- is programmed into the supply-side.

@@

Homophily is central to market research products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].[^4] Classification situations engender markets where a vertical hierarchy of products are matched to a vertical hierarchy of consumers. We observe similar processes in our case study. @RN4982 categorizes students into four market segments -- local, in-state, regional, and national -- based on their score sending behavior and then evaluates the attractiveness of each Geomarket based on how many "regional" and "national" students live there.
<!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables, @RN5005 we cannot show the effect of quantification. However, this study provides inisight about the mechanism underlying the effect. By analyzing and simulating *Student Search Service* purchases that filter on Geomarkets we show how Geomarkets reproduce historical race-based inequality in college access.

@@

**Creating Geomarkets and the Market Segment Model.** In 1978, University of Pennsylvania professor Robert Zemsky, was tasked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s. <!--The geographic concepts inscribed in the Market Segment Model were based on conversations with college admissions officers. -->  The "initial task was to define enrollment markets in a manner consistent with admissions officers' intuitive understanding of student pools" [@RN4982,p. 4]. Quoting an admissions officer, @RN4982 [p. 11] write, "'There are only three kinds of college-bound students: those who want to live at home, those who want to live on campus but bring their laundry home, and those who want to go far enough from home that Mom and Dad can't visit without calling first.'" As such, [@RN4982] created three geographic [@RN4982, p. 11], region, state, and community. The three regions were New England, Middle States, and the South. They "divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" (p. 11). These enrollment markets, later called Geomarkets, were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. @RN4982[pp. 11-12] briefly describes the creation of Geomarket borders:

@@

The goal of the Market Segment Model was to predict how student demand for a particular college varies by Geomarket, based on the characteristics of households in the Geomarket. Zemsky reached out to College Board because "we needed a database that described most institutions and most students" [@RN4982, p. x].  <!--Zemsky reached out to College Board because "we needed a database that described most institutions and most students\ldots Coincidentally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential" [@RN4982, p. x]. -->In 1979, College Board began providing funding and data on the score-sending behavior of SAT test-takers.
